---
title: "What the confidence criteria actually control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What the confidence criteria actually control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spawnR)
```

A parentage assignment can fail in two directions. You can reject a parent who
really is the parent, or you can accept a candidate who is not. The two
threshold criteria in this package control one of those each, and neither
controls both. Which one you use changes what your assignment rate means, and
the difference gets larger as the candidate pool grows.

## The two rules

**The trio-specific rule** of Amiri Roudbar et al. (2025), `criterion =
"pairwise"`, simulates the LOD score a *true* parent of this particular
offspring would produce, given its genotypes and its pattern of missing data,
and takes a lower-tail quantile. A candidate is accepted when its score is one a
true parent would plausibly have produced. What that bounds is the chance of
rejecting a true parent. It says nothing about how many unrelated candidates
also clear the bar.

**The population rule** of Marshall et al. (1998), `criterion = "delta"`,
defines confidence as the proportion of simulated assignments above a threshold
in which the top candidate really is the parent, and simulates the whole
candidate pool including the chance that the true parent was never sampled. What
that bounds is the share of assignments that are wrong.

## A factorial simulation

Fifteen microsatellite loci with ten alleles each, 1% genotyping error, 5%
missing data, 400 offspring. Candidate pools of 50 to 5000 males, and a
proportion `prop_sampled` of offspring whose true sire is actually in the pool.
The full script is in `system.file("studies", "01-criteria-factorial.R", package
= "spawnR")`.

Three numbers per cell. **Assigned** is the share of offspring given a parent.
**Precision** is the share of those assignments that are correct. **Recall** is
the share of offspring *whose sire was in the pool* that were correctly
assigned.

```{r, echo = FALSE}
res <- read.csv(text = "
n_cand,prop_sampled,pw_assigned,pw_precision,pw_recall,d_assigned,d_precision,d_recall
50,1.00,97.2,100.0,97.2,99.8,100.0,99.8
50,0.80,83.0,99.4,97.3,86.0,98.5,100.0
50,0.50,48.5,95.4,96.9,52.5,90.5,99.5
200,1.00,97.2,99.7,97.0,100.0,99.8,99.8
200,0.80,82.0,98.5,96.7,86.5,96.0,99.4
200,0.50,54.2,93.5,96.7,54.0,93.1,95.7
1000,1.00,97.8,98.2,96.0,100.0,97.8,97.8
1000,0.80,82.5,91.8,95.3,77.0,94.5,91.5
1000,0.50,59.8,75.3,96.3,40.8,92.0,80.2
5000,1.00,98.5,93.1,91.8,94.5,95.0,89.8
5000,0.80,91.5,83.1,92.7,64.0,94.1,73.5
5000,0.50,77.8,54.0,91.8,26.0,95.2,54.1")
knitr::kable(res, col.names = c("candidates", "prop sampled",
  "assigned", "precision", "recall", "assigned", "precision", "recall"),
  caption = "Left three columns: pairwise criterion at alpha = 0.01. Right three: Delta at 95%.")
```

## The two rules fail in opposite directions

Read down the **pairwise recall** column: 97, 97, 97, 97, 97, 97, 96, 95, 96,
92, 93, 92. It barely moves. Pool size and sampling fraction change it by a few
points at most, which is exactly right, because the criterion is built from the
true-parent distribution and that distribution does not know how many other
candidates exist.

Now read **pairwise precision**: 100, 99, 95, 100, 99, 94, 98, 92, 75, 93, 83,
54. It collapses. Two things drive it and they multiply. A bigger pool offers
more unrelated males, one of which may clear a fixed bar by chance. A lower
sampling fraction means more offspring have no true sire available at all, so
every assignment made for them is necessarily wrong.

The Delta criterion is the mirror image. Its precision holds between 90% and
100% everywhere, which is what it was designed to do. Its recall is what gives
way: 99.8 down to 54.1. It buys precision by refusing to assign, and in the
hardest cell it assigns only 26% of offspring.

So the choice is not which criterion is better. It is which error you would
rather make, and how you intend to describe the result.

## What this means for a results section

If you report "we assigned parentage to 78% of offspring with 99% confidence"
using the pairwise criterion in a 5000-candidate pool with half the sires
sampled, that 99% refers to the chance of not rejecting a true sire. Nearly half
of the assignments you just reported are wrong. The number is not incorrect; it
is answering a question nobody asked.

If you report the same study under Delta, you assign 26% of offspring and about
95% of those are right, but you have thrown away most of your data and the
discarded offspring are not a random subset of them.

Neither rule reports the thing you would most like to state, which is the
probability that *this* assignment is correct. That is what
`vignette("posterior", package = "spawnR")` is about.

## Seeing it on a small scale

```{r}
sim <- simulate_population(n = 400, n_loci = 15, n_alleles = 10,
                           n_offspring = 60, error = 0.01, seed = 11)
p <- allele_freqs(sim$genotypes)
ped <- sim$pedigree

pw <- assign_parentage(sim$genotypes, ped$offspring, sim$sires, p,
                       error = 0.01, criterion = "pairwise", nsim = 1000)
truth <- ped$sire[match(pw$offspring, ped$offspring)]
c(assigned = sum(pw$assigned),
  correct  = sum(pw$assigned & pw$candidate == truth))
```

With 200 candidates and every sire sampled this is the easy corner of the table,
and the pairwise rule does well. The failure only appears once the pool is large
or the sampling fraction is low, which is why it is easy to miss in a pilot
study and costly in a real one.

## References

Marshall, T.C., Slate, J., Kruuk, L.E.B. & Pemberton, J.M. (1998) Statistical
confidence for likelihood-based paternity inference in natural populations.
*Molecular Ecology* **7**, 639-655.

Kalinowski, S.T., Taper, M.L. & Marshall, T.C. (2007) Revising how the computer
program CERVUS accommodates genotyping error increases success in paternity
assignment. *Molecular Ecology* **16**, 1099-1106.

Amiri Roudbar, M., Mousavi, S.F., Akbarzadeh, M., Brounts, S.H. & Momen, M.
(2025) Pairwise paternity assignment with forward-backward simulations.
*Ecology and Evolution* **15**(10), e72230.
