---
title: "Posterior assignment and false discovery rate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior assignment and false discovery rate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spawnR)
```

The companion vignette shows that the two threshold criteria fail in opposite
directions: the trio-specific rule holds recall steady and loses precision as
the candidate pool grows, the population rule holds precision steady and loses
recall. That trade is not a law of nature. It is a consequence of using a
*decision rule* when what you want is a *probability about one assignment*.

## The model

For one offspring with `n` candidates, either the true parent is among them,
with prior probability `prop_sampled`, or it is not. Given that it is, no
candidate is favoured a priori, so

$$\Pr(H_j) = \pi/n, \qquad \Pr(H_0) = 1 - \pi.$$

The genotypes of the candidates you are *not* testing appear identically under
every hypothesis and cancel, which leaves the likelihood ratio of candidate $j$
against "the parent was never sampled" equal to $e^{L_j}$, where $L_j$ is
exactly the LOD score `parentage_lod()` already returns. So

$$\Pr(H_j \mid g) = \frac{(\pi/n)\,e^{L_j}}{(1-\pi) + (\pi/n)\sum_k e^{L_k}}.$$

Two features matter. The sum in the denominator runs over the whole pool, so
adding candidates dilutes the prior and the same LOD buys less confidence in a
bigger pool. That is the pool-size problem, solved by construction rather than
by a correction. And the $(1-\pi)$ term gives the model somewhere to put its
probability when no candidate is convincing, reported as `post_unsampled`.

Everything is computed on the log scale, so LOD scores of any size are safe.

## It is calibrated

Same factorial design as the other vignette: pools of 50 to 5000, sampling
fractions of 1, 0.8 and 0.5, 500 offspring per cell, 15 microsatellites.
Pooling all cells and binning by the stated posterior:

```{r, echo = FALSE}
cal <- read.csv(text = "
bin,n,stated,observed
'[0, 0.5]',1509,0.073,0.089
'(0.5, 0.8]',272,0.651,0.614
'(0.8, 0.9]',165,0.854,0.855
'(0.9, 0.95]',177,0.928,0.938
'(0.95, 0.99]',465,0.976,0.981
'(0.99, 0.999]',826,0.996,0.999
'(0.999, 1]',2586,1.000,1.000")
knitr::kable(cal, col.names = c("posterior bin", "n", "mean stated", "observed correct"),
             caption = "Stated probability against observed frequency of being right.")
```

The stated probability is the observed frequency, across the whole range. A
posterior of 0.85 means about 85% of such assignments are right, and a posterior
of 0.65 means about 65%, which is the property that lets you use the number
directly rather than as a ranking.

## Precision no longer depends on the pool

```{r, echo = FALSE}
pr <- read.csv(text = "
n_cand,prop_sampled,assigned,precision,recall
50,1.0,98.8,100.0,98.8
50,0.8,75.8,100.0,94.5
50,0.5,48.6,99.6,91.7
200,1.0,96.6,100.0,96.6
200,0.8,72.4,99.7,89.8
200,0.5,40.8,100.0,81.0
1000,1.0,89.4,100.0,89.4
1000,0.8,64.6,99.4,79.5
1000,0.5,35.8,99.4,73.9
5000,1.0,75.8,99.7,75.6
5000,0.8,50.6,99.2,64.9
5000,0.5,26.2,98.5,51.6")
knitr::kable(pr, col.names = c("candidates", "prop sampled", "assigned %",
                               "precision %", "recall %"),
             caption = "Accepting assignments at posterior >= 0.95.")
```

Precision runs between 98.5% and 100% in every cell. Compare the pairwise
criterion over the same design, which fell to 54%.

Recall still declines with pool size and sampling fraction, and it should. With
5000 candidates and half the sires never sampled, the data genuinely do not
identify most sires. The difference is that the posterior *says so* instead of
assigning anyway. The offspring you lose are reported as uncertain rather than
silently mis-assigned.

## How wrong can prop_sampled be?

It is a prior, so it is fair to ask what happens when you get it wrong.
Precision at posterior >= 0.95, with the prior correctly specified, fixed at
0.9, and fixed at 0.5:

```{r, echo = FALSE}
sens <- read.csv(text = "
n_cand,true_pi,correct,assumed_0.9,assumed_0.5
50,1.0,100.0,100.0,100.0
50,0.5,99.6,98.5,99.6
200,1.0,100.0,100.0,100.0
200,0.5,100.0,97.4,100.0
1000,1.0,100.0,100.0,100.0
1000,0.5,99.4,98.5,99.4
5000,1.0,99.7,100.0,100.0
5000,0.5,98.5,96.2,98.5")
knitr::kable(sens, col.names = c("candidates", "true prop sampled",
  "prior correct", "prior set 0.9", "prior set 0.5"),
  caption = "Precision (%) at posterior >= 0.95 under a misspecified prior.")
```

The worst cell costs 2.3 points of precision for a prior that was wrong by 0.4.
Wherever the posterior is high the likelihood dominates, so the prior mostly
moves borderline cases. Still, report which value you used, and vary it.

One value to avoid is exactly 1. It asserts that the parent is certainly in the
pool, which removes the model's ability to answer "none of these" and forces a
best guess for every offspring.

## False discovery rate

Because the posteriors are calibrated, the expected number of wrong assignments
in any accepted set is just the sum of `1 - posterior` over that set. No extra
simulation is needed.

```{r}
sim <- simulate_population(n = 500, n_loci = 15, n_alleles = 10,
                           n_offspring = 100, error = 0.01,
                           prop_missing = 0.05, seed = 4)
p <- allele_freqs(sim$genotypes)
ped <- sim$pedigree

m <- parentage_lod(sim$genotypes, ped$offspring, sim$sires, p,
                   error = 0.01, output = "matrix")
post <- parentage_posterior(m, prop_sampled = 0.9)
parentage_fdr(post, target = 0.05)
```

`parentage_fdr()` accepts offspring in decreasing order of posterior and stops
when the expected error rate reaches the target, returning the largest set whose
expected FDR is at most `target`. Checked against the truth, the realised rate
across the factorial ran between 0% and 5.5% against a 5% target.

That is the practical payoff. The quantity being controlled is the one a results
section reports, and it does not drift when the pool changes size.

```{r}
truth <- ped$sire[match(post$offspring, ped$offspring)]
cut <- parentage_fdr(post, target = 0.05)$cutoff
sel <- post$posterior >= cut
c(assigned = sum(sel),
  wrong = sum(post$candidate[sel] != truth[sel]),
  expected_wrong = round(sum(1 - post$posterior[sel]), 2))
```

## Two things it does not fix

**Relatives in the pool.** The prior treats candidates as exchangeable and the
likelihood treats them as unrelated to the offspring except under the hypothesis
being tested. A full sib of the true sire violates both, and the posterior will
be too confident about whichever relative scores highest. This is the same
limitation CERVUS addresses with its relatedness options, and it is not
addressed here.

**A wrong error rate.** The LOD scores are only as good as `error`. Estimate it
from known parent-offspring pairs with `estimate_error_rates()` rather than
assuming a round number.

## References

Marshall, T.C., Slate, J., Kruuk, L.E.B. & Pemberton, J.M. (1998) Statistical
confidence for likelihood-based paternity inference in natural populations.
*Molecular Ecology* **7**, 639-655.

Kalinowski, S.T., Taper, M.L. & Marshall, T.C. (2007) Revising how the computer
program CERVUS accommodates genotyping error increases success in paternity
assignment. *Molecular Ecology* **16**, 1099-1106.

Amiri Roudbar, M., Mousavi, S.F., Akbarzadeh, M., Brounts, S.H. & Momen, M.
(2025) Pairwise paternity assignment with forward-backward simulations.
*Ecology and Evolution* **15**(10), e72230.
