---
title: "Relatives in the candidate pool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Relatives in the candidate pool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spawnR)
```

The posterior in `vignette("posterior")` assumes candidates are unrelated to the
offspring except under the hypothesis being tested. Real candidate pools are
rarely like that: hatchery broodstock, a spawning shoal, a trapped run, all
contain sibs and cousins. This vignette measures what that does, shows a
plausible correction that does not work and why, and gives one that does.

## It takes two things going wrong at once

Pools were seeded with known full sibs of the true sire, then the sire himself
was withheld from a fraction of offspring. Assignments taken at posterior >= 0.95.

```{r, echo = FALSE}
d <- read.csv(text = "
panel,full_sibs,pi_sire,stated,observed
15 msat,0,1.0,1.000,1.000
15 msat,0,0.5,0.996,0.997
15 msat,10,1.0,0.996,0.997
15 msat,3,0.5,0.996,0.921
15 msat,10,0.5,0.993,0.838
96 SNP,10,1.0,0.999,0.999
96 SNP,3,0.5,0.997,0.967
96 SNP,10,0.5,0.997,0.921")
knitr::kable(d, col.names = c("panel", "full sibs of sire", "prop sire sampled",
                              "stated posterior", "observed correct"))
```

Neither condition alone does harm. Ten full sibs with every sire sampled: 0.996
stated, 0.997 observed. Half the sires missing with no relatives: 0.996 stated,
0.997 observed. Both together: 0.993 stated, **0.838** observed.

The reason is that the model has only two explanations for a candidate who
shares a lot of alleles with the offspring, namely *parent* and *stranger*. When
the parent is genuinely absent, his full sib is the best available explanation,
and with no hypothesis for "close relative of the parent" the probability that
should have gone to `post_unsampled` lands on the sib instead. With the sire
present he simply outscores his sib and nothing goes wrong, which is why this
failure is invisible in a validation set where every parent is known.

## A correction that does not work

The obvious repair is to make the alternative hypothesis a mixture: a non-parent
candidate is unrelated with probability 0.95 and avuncular with probability
0.05, so the per-locus denominator becomes

$$k_0 P(o) + k_1 T(o \mid c) + k_2 \mathbb{1}\{o = c\}$$

evaluated at the mixture-averaged coefficients. The likelihood is linear in
$(k_0, k_1, k_2)$, so a mixture of classes really is the average vector, and the
algebra is exact. It also barely helps:

```{r, echo = FALSE}
d2 <- read.csv(text = "
cell,naive,product_of_mixtures,mixture_of_products
15 msat / 3 FS / pi 0.5,0.921,0.925,0.978
15 msat / 10 FS / pi 0.5,0.838,0.848,0.907
96 SNP / 3 FS / pi 0.5,0.967,0.967,0.991
96 SNP / 10 FS / pi 0.5,0.921,0.926,0.987")
knitr::kable(d2, col.names = c("cell", "naive", "per-locus mixture",
                               "multilocus mixture"),
             caption = "Observed proportion correct among assignments at posterior >= 0.95.")
```

The averaged coefficient is small, $k_1 = 0.025$ for a pool that is 5%
avuncular, so each locus gains a modest constant to its denominator. True
parents and their sibs both lose about the same amount, and the *difference*
between them, which is the only thing that decides the assignment, is almost
unchanged.

The deeper error is structural. A candidate either is or is not a sib, one state
for the whole genome. Mixing inside the per-locus denominator instead models an
independent coin flip at every locus, which is a different and much weaker
hypothesis. It is a **product of mixtures** where the biology is a **mixture of
products**.

## The correction that does work

Evaluate the multilocus likelihood separately under each relationship class and
mix at the end. Writing $L_c$ for the LOD against class $c$ alone,

$$\mathrm{LOD} = -\log \sum_c w_c\, e^{-L_c}.$$

This is dominated by whichever alternative explains the pair best, so a
candidate who looks like a sib is scored against the sib hypothesis rather than
against a stranger. That is the `relatedness` argument:

```{r}
sim <- simulate_population(n = 300, n_loci = 15, n_alleles = 10,
                           n_offspring = 40, error = 0.01, seed = 7)
p <- allele_freqs(sim$genotypes)
ped <- sim$pedigree

naive <- parentage_lod(sim$genotypes, ped$offspring, sim$sires, p, error = 0.01)
aware <- parentage_lod(sim$genotypes, ped$offspring, sim$sires, p, error = 0.01,
                       relatedness = c(unrelated = 0.95, avuncular = 0.05))

kt <- paste(ped$offspring, ped$sire)
c(naive = mean(naive$lod[paste(naive$offspring, naive$candidate) %in% kt]),
  aware = mean(aware$lod[paste(aware$offspring, aware$candidate) %in% kt]))
```

True sires score lower, as they should: against a pool that might contain their
sibs, the evidence that *this* male rather than a brother is the parent is
genuinely weaker.

Classes are named for the **candidate-to-offspring** relationship, which is one
step further out than people usually think. A full sib of the true father is the
offspring's uncle, so the class is `avuncular`, not `full_sib`. See
`?ibd_mixture`.

```{r}
ibd_mixture(unrelated = 0.95, avuncular = 0.05)
```

## What it costs, and what it does not fix

Recall falls. In the hardest cell, 15 microsatellites with ten full sibs and
half the sires unsampled, the assignment rate at posterior >= 0.95 drops from
47% to 7%. That is the honest answer: with those markers and that pool you
cannot separate a sire from ten of his brothers, and the model now says so
instead of guessing.

That same cell is also the one the fix does not fully repair, reaching 0.907
observed against 0.963 stated. Everything else recalibrates to within a point or
two. So treat the correction as substantially reducing the bias, not abolishing
it, and treat very high relatedness with weak markers as a study-design problem
rather than an analysis problem.

Two further limits. The correction is not implemented for
`type = "pair"`. And it needs you to state the pool composition; there is a
screen to tell you whether it matters,

```{r}
pool_relatedness(sim$genotypes, ped$offspring[1:4], sim$sires, p,
                 error = 0.01, nsim = 400)
```

but it reports a lower bound, not an estimate. In a pool seeded with ten full
sibs of the sire plus the sire, 5.5% by construction, it returns about 3%,
because only relatives extreme enough to clear the threshold are counted. Use it
to decide whether to reach for `relatedness`, and set the weights from how the
pool was actually collected.

## References

Marshall, T.C., Slate, J., Kruuk, L.E.B. & Pemberton, J.M. (1998) *Molecular
Ecology* **7**, 639-655.

Kalinowski, S.T., Taper, M.L. & Marshall, T.C. (2007) *Molecular Ecology*
**16**, 1099-1106.

Amiri Roudbar, M., Mousavi, S.F., Akbarzadeh, M., Brounts, S.H. & Momen, M.
(2025) *Ecology and Evolution* **15**(10), e72230.
