---
title: "Replication and external validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replication and external validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(spawnR)
```

This package was written clean-room, from published equations and prose, without
reading any existing implementation. That buys independence but costs the usual
reassurance, and it raises a specific worry: the genotyping-error coefficients
here were **re-derived** rather than transcribed, because the ones printed in the
sources do not sum to one and so cannot be a probability decomposition.

There is a reason to think the re-derivation is recoverable rather than
invented. The corrigendum to Kalinowski et al. (2007) states that the typesetting
errors in that paper's appendix "do not affect any version of the Cervus
software." CERVUS therefore implements the intended equations, and the printed
ones were damaged on the way to the page. Two consequences follow, and both are
testable.

## A published number, reproduced

Kalinowski et al. (2007) state their simulation parameters exactly, which makes
their headline result a falsifiable target:

> "We assumed that allele frequencies in the population were [0.25, 0.25, 0.2,
> 0.15, 0.05, 0.05, 0.02, 0.01, 0.01, 0.005, 0.005]." Each iteration "began by
> randomly selecting actual genotypes for 100 unrelated adult males and 100
> unrelated adult females", with "a genotyping error rate of 0.01", over "one
> hundred thousand simulations".

and report that with six loci, at a Δ 0.99 confidence level, "the corresponding
result for the corrected equations was 73%".

Running that specification through this package, with nothing taken from CERVUS:

```{r, echo = FALSE}
k <- read.csv(text = "
loci,crit99,assigned,precision
4,4.438,25.3,99.0
5,4.260,46.3,99.0
6,3.440,72.1,99.0
7,1.380,89.4,99.0
8,0.694,96.4,99.0")
knitr::kable(k, col.names = c("loci", "critical Delta (99%)", "assigned %",
                              "precision %"),
             caption = "40,000 simulations per row. Kalinowski et al. report 73% for six loci.")
```

**72.1% against a reported 73%.** The row is not fitted to anything; the allele
spectrum, locus count, candidate count and error rate all come from the paper,
and the equations come from the re-derivation. Precision sits at exactly 99.0%
at every locus count, which is the Δ 0.99 criterion doing what it claims.

The script is at
`system.file("studies", "06-kalinowski-replication.R", package = "spawnR")`.

### What cannot be reproduced

The same passage reports **51%** for Marshall et al.'s *original* equations, and
that figure is out of reach here. Reproducing it needs Marshall's original error
formula as printed, and that equation could not be obtained: the 1998 paper is
closed-access with no open copy, and the only description available is
Kalinowski's characterisation of the mistake, that Marshall "mistakenly indicated
that the probability of making a genotyping error and observing g is equal to ε,
instead of εP(g)". That is enough to know what went wrong and not enough to
reconstruct the printed expression, which does not factor the way the corrected
one does. Rather than guess at it and present the guess as a replication, this
package implements only the corrected form.

## Comparing directly against CERVUS

The stronger test is running both programs on the same genotypes and diffing the
LOD scores pair by pair. CERVUS is Windows software, so this is something you run
rather than something the package can do for you, but the import and comparison
are provided.

```r
# 1. In CERVUS: run your analysis and export the parentage results to CSV.
# 2. In R, read the same genotypes and the same allele frequencies.
g <- genotypes(read.csv("genotypes.csv"), id_col = "id")
p <- allele_freqs(g, min_freq = 0)          # see "getting the inputs to match"

mine <- parentage_lod(g, offspring, candidates, p, error = 0.01,
                      type = "both_unknown")     # pair LOD
theirs <- read_cervus_results("cervus_parentage.csv", which = "pair")

compare_cervus(mine, theirs)
```

`read_cervus_results()` detects columns by pattern rather than assuming a layout,
because CERVUS's header text varies by analysis type and version and no
authoritative list is published. It prints the mapping it used, so check that
against your file instead of trusting it.

`compare_cervus()` reports how many rows matched, the correlation, the mean and
maximum absolute difference, how many rows exceed a tolerance, and whether the
mismatch and loci-compared counts agree. Agreement on those integer counts is
worth as much as the LOD correlation: it confirms both programs are reading the
same genotypes the same way, which is the usual cause of a confusing diff.

### Getting the inputs to match

Four things have to line up before a difference means anything.

**Allele frequencies.** `allele_freqs()` floors every frequency at `1/(2n)` by
default, the frequency of one unobserved allele copy, which keeps the likelihood
finite when a candidate carries an allele never seen in the reference sample.
CERVUS does not do this. Pass `min_freq = 0` to match, and check CERVUS's own
minimum allele frequency setting.

**One error rate.** CERVUS assumes a single rate across loci, so pass a scalar
`error`, not the per-locus vector from [estimate_error_rates()].

**The same configuration.** A CERVUS *pair* LOD corresponds to
`type = "both_unknown"`; a *trio* LOD, with the other parent supplied,
corresponds to `type = "one_known"`. Comparing a trio against a pair produces a
large, meaningless difference.

**No relatedness correction.** Leave `relatedness` and `ibd` unset. CERVUS's
alternative hypothesis is an unrelated candidate; a relatedness-aware score is
deliberately a ratio against something else and is not expected to match.

### Reading a disagreement

A correlation near 1 with a small constant offset usually means the allele
frequencies differ slightly, often the `min_freq` flooring above. Scattered
large differences on specific pairs usually mean missing data is being read
differently, which the loci-compared column will show. Differences that scale
with the number of mismatching loci point at the error model itself, which is the
case worth telling me about.

Two things will *not* match and are not errors. Confidence thresholds from
[delta_critical()] come from this package's own simulation with its own random
draws, so they differ from CERVUS's run to run even when the method agrees. And
anything from [parentage_posterior()] has no CERVUS counterpart at all.

## References

Marshall, T.C., Slate, J., Kruuk, L.E.B. & Pemberton, J.M. (1998) Statistical
confidence for likelihood-based paternity inference in natural populations.
*Molecular Ecology* **7**, 639-655.

Kalinowski, S.T., Taper, M.L. & Marshall, T.C. (2007) Revising how the computer
program CERVUS accommodates genotyping error increases success in paternity
assignment. *Molecular Ecology* **16**, 1099-1106. Corrigendum: *Molecular
Ecology* **19**, 1512.

Amiri Roudbar, M., Mousavi, S.F., Akbarzadeh, M., Brounts, S.H. & Momen, M.
(2025) Pairwise paternity assignment with forward-backward simulations.
*Ecology and Evolution* **15**(10), e72230.
